---
title: "index"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message=FALSE, results='hide')
library(dplyr)
library(tidyr)
library(plotly)
library(ggplot2)
library(maps)
options(dplyr.summarise.inform = FALSE)
options(dplyr.select.inform = FALSE)
```
## Introduction & Summary information

### Dataset introduction
I have used two dataset, Incarceration_trends and state code mapping dataset. The Incarceration_trends dataset includes the incarceration data and provides county-level data on prison and jail incarceration for the entire United States. The data also contains prison data grouped by different races. State code dataset includes the mapping of state region and state code (e.g. California -> CA).

### Problems
In this report, I would like to solve several problems, including:

1. As time goes by, is there a decrements of prison ratio?

2. Is there noteworthy prison ratio difference between different races in different level of urbanity?

3. As time goes by, is there noteworthy prison ratio difference between different races?

### Variables
I have used several variables from the dataset and generate several variables from them.

1. Year, range: from 1970 to 2018

2. State, the State code of the data. (e.g. CA)

3. Urbanicity, the urbanicity of the data, have 4 unique values: "rural", "small/mid", "suburban", "urban".

4. total_pop_15to64, the population count of ages 15 to 64, range: from 0 to 6958134

5. {races}_pop_15to64, the population of people between 15 and 64 with 5 different races: white, native, black, latinx, aapi.

6. total_jail_pop, total jail population count, range: from 0 to 23467

7. {races}_jail_pop, jail population with 5 difference races: white, native, black, latinx, aapi.

8. total_prison_ratio = total_jail_pop / total_pop_15to64

9. {races}_prison_ratio = {races}_jail_pop / {races}_pop_15to64

### Data preprocessing and cleaning
Because the calculation of prison ration relies on the population data of different races, which are N/A for the data before 1990, so I remove the data before 1990. And replace the rest N/A values with 0.

```{r, echo=FALSE}
prison_csv <- read.csv(file = "source/incarceration_trends.csv", header = T)

prison_data <- prison_csv %>%
  select(
    year,
    state,
    total_pop_15to64,
    aapi_pop_15to64,
    black_pop_15to64,
    latinx_pop_15to64,
    native_pop_15to64,
    white_pop_15to64,
    urbanicity,
    total_jail_pop,
    aapi_jail_pop,
    black_jail_pop,
    latinx_jail_pop,
    native_jail_pop,
    white_jail_pop
  ) %>%
  replace_na(
    list(
      total_pop_15to64 = 0,
      aapi_pop_15to64 = 0,
      black_pop_15to64 = 0,
      latinx_pop_15to64 = 0,
      native_pop_15to64 = 0,
      white_pop_15to64 = 0,
      total_jail_pop = 0,
      aapi_jail_pop = 0,
      black_jail_pop = 0,
      latinx_jail_pop = 0,
      native_jail_pop = 0,
      white_jail_pop = 0
    )
  ) %>%
  filter(year >= 1990)
```

## Time trend Chart
For problem 1, I have plot the time trend of prison ratio of the top 10 population states in 2018. The plot shows:

1. The prison ratio of state PA keeps increasing from 1990 to 2018.

2. There is a decrements of prison ratio of FL between 2007 and 2018.

3. There are slightly increase of average prison ratio for those states.

```{r, echo=FALSE}
top_10_pop_state <- prison_data %>%
  filter(year == max(year)) %>%
  group_by(state) %>%
  summarise(total_pop_15to64 = sum(total_pop_15to64)) %>%
  arrange(desc(total_pop_15to64)) %>%
  top_n(10) %>%
  pull(state)

top_10_state_pri_rate <- prison_data %>%
  filter(state %in% top_10_pop_state) %>%
  group_by(year, state) %>% 
  summarise(
    total_jail_pop = sum(total_jail_pop),
    total_pop_15to64 = sum(total_pop_15to64)
  ) %>%
  mutate(prison_rate = total_jail_pop / total_pop_15to64)

# figure 1
p <- ggplot(data=top_10_state_pri_rate, aes(x=year, y=prison_rate, color=state)) +
  geom_line() + 
  geom_point() +
  labs(title="Population top 10 states prison rate by year", x="Year", y="Prison rate")
p

```

For problem 3, I have plot the time trend of prison ratio by different races for the state with highest prison ratio. From the plot we can find out that the black has higher prison ratio than other races, and the prison ratio has a decrements between 2009 and 2018. So I think the inequality do exists but the problem has been mitigated since recent years.

```{r, echo=FALSE}
jail_rate_by_state <- prison_data %>%
  filter(year >= 2000) %>%
  group_by(year, state) %>%
  summarise(
    total_pop_15to64 = sum(total_pop_15to64),
    total_jail_pop = sum(total_jail_pop)
  ) %>%
  mutate(
    total_rate = total_jail_pop / total_pop_15to64
  ) %>%
  group_by(state) %>%
  summarise(
    total_rate = mean(total_rate)
  )

state_top_jail_rate <- jail_rate_by_state %>%
  arrange(desc(total_rate)) %>%
  top_n(1) %>%
  pull(state)

jail_rate_race_year <- prison_data %>%
  filter(year >= 2000, state == state_top_jail_rate) %>%
  group_by(year, state) %>%
  summarise(
    total_pop_15to64 = sum(total_pop_15to64),
    aapi_pop_15to64 = sum(aapi_pop_15to64),
    black_pop_15to64 = sum(black_pop_15to64),
    latinx_pop_15to64 = sum(latinx_pop_15to64),
    native_pop_15to64 = sum(native_pop_15to64),
    white_pop_15to64 = sum(white_pop_15to64),
    total_jail_pop = sum(total_jail_pop),
    aapi_jail_pop = sum(aapi_jail_pop),
    black_jail_pop = sum(black_jail_pop),
    latinx_jail_pop = sum(latinx_jail_pop),
    native_jail_pop = sum(native_jail_pop),
    white_jail_pop = sum(white_jail_pop)
  ) %>%
  mutate(
    total_rate = total_jail_pop / total_pop_15to64,
    aapi_rate = aapi_jail_pop / aapi_pop_15to64,
    black_rate = black_jail_pop / black_pop_15to64,
    latinx_rate = latinx_jail_pop / latinx_pop_15to64,
    native_rate = native_jail_pop / native_pop_15to64,
    white_rate = white_jail_pop / white_pop_15to64
  )


aapi_jail_rate_by_state <- jail_rate_race_year %>%
  mutate(
    rate = aapi_rate,
    type = "aapi"
  ) %>%
  select(year, rate, type)

black_jail_rate_by_state <- jail_rate_race_year %>%
  mutate(
    rate = black_rate,
    type = "black"
  ) %>%
  select(year, rate, type)

latinx_jail_rate_by_state <- jail_rate_race_year %>%
  mutate(
    rate = latinx_rate,
    type = "latinx"
  ) %>%
  select(year, rate, type)

native_jail_rate_by_state <- jail_rate_race_year %>%
  mutate(
    rate = native_rate,
    type = "native"
  ) %>%
  select(year, rate, type)

white_jail_rate_by_state <- jail_rate_race_year %>%
  mutate(
    rate = white_rate,
    type = "white"
  ) %>%
  select(year, rate, type)

jail_rate_race_year <- aapi_jail_rate_by_state %>%
  bind_rows(black_jail_rate_by_state) %>%
  bind_rows(latinx_jail_rate_by_state) %>%
  bind_rows(white_jail_rate_by_state) %>%
  bind_rows(native_jail_rate_by_state) 

# fig 4

p <- ggplot(data=jail_rate_race_year, aes(x=year, y=rate, color=type)) +
  geom_line() + 
  geom_point() +
  labs(title="Prison rate top 1 state (LA) by time", x="Year", y="Prison rate")
p

```

## Variable Comparison Chart
For problem 2, I compared the prison ratio between urbanicity and races. I found that the black have higher prison ratio than other races, and the people lived in rural have higher prison ratio. So I think the inequality do exists.

``` {r, echo=FALSE}

jail_rate_by_urbanicity <- prison_data %>%
  filter(urbanicity != "") %>%
  group_by(urbanicity, year) %>%
  summarise(
    total_pop_15to64 = sum(total_pop_15to64),
    aapi_pop_15to64 = sum(aapi_pop_15to64),
    black_pop_15to64 = sum(black_pop_15to64),
    latinx_pop_15to64 = sum(latinx_pop_15to64),
    native_pop_15to64 = sum(native_pop_15to64),
    white_pop_15to64 = sum(white_pop_15to64),
    total_jail_pop = sum(total_jail_pop),
    aapi_jail_pop = sum(aapi_jail_pop),
    black_jail_pop = sum(black_jail_pop),
    latinx_jail_pop = sum(latinx_jail_pop),
    native_jail_pop = sum(native_jail_pop),
    white_jail_pop = sum(white_jail_pop)
  ) %>%
  mutate(
    total_rate = total_jail_pop / total_pop_15to64,
    aapi_rate = aapi_jail_pop / aapi_pop_15to64,
    black_rate = black_jail_pop / black_pop_15to64,
    latinx_rate = latinx_jail_pop / latinx_pop_15to64,
    native_rate = native_jail_pop / native_pop_15to64,
    white_rate = white_jail_pop / white_pop_15to64
  ) %>%
  group_by(urbanicity) %>%
  summarise(
    total_rate = mean(total_rate),
    aapi_rate = mean(aapi_rate),
    black_rate = mean(black_rate),
    latinx_rate = mean(latinx_rate),
    native_rate = mean(native_rate),
    white_rate = mean(white_rate)
  )

aapi_jail_rate_by_state <- jail_rate_by_urbanicity %>%
  mutate(
    rate = aapi_rate,
    type = "aapi"
  ) %>%
  select(urbanicity, rate, type)

black_jail_rate_by_state <- jail_rate_by_urbanicity %>%
  mutate(
    rate = black_rate,
    type = "black"
  ) %>%
  select(urbanicity, rate, type)

latinx_jail_rate_by_state <- jail_rate_by_urbanicity %>%
  mutate(
    rate = latinx_rate,
    type = "latinx"
  ) %>%
  select(urbanicity, rate, type)

native_jail_rate_by_state <- jail_rate_by_urbanicity %>%
  mutate(
    rate = native_rate,
    type = "native"
  ) %>%
  select(urbanicity, rate, type)

white_jail_rate_by_state <- jail_rate_by_urbanicity %>%
  mutate(
    rate = white_rate,
    type = "white"
  ) %>%
  select(urbanicity, rate, type)

jail_rate_by_urbanicity <- aapi_jail_rate_by_state %>%
  bind_rows(black_jail_rate_by_state) %>%
  bind_rows(latinx_jail_rate_by_state) %>%
  bind_rows(white_jail_rate_by_state) %>%
  bind_rows(native_jail_rate_by_state) 
# fig 2 
g <- ggplot(data=jail_rate_by_urbanicity, aes(x=type, y=rate, fill=urbanicity)) +
  geom_bar(stat="identity", position=position_dodge(), width=0.5) + 
  labs(title="Jail rate by race", x="Race", y="Jail rate")
g


```

## Map
I have plot the map of US with the average prison ratio from 2000 to 2018, because I think prison ratio can reflect the public security situation of states and people can check and compare that in a single map.

```{r, echo=FALSE}
MainStates <- map_data("state")
StateToCode <- read.csv(file = "source/csvData.csv", header = T)
StateToCode$region = tolower(StateToCode$State)

MainStates <- MainStates %>%
  left_join(StateToCode, by = c("region" = "region"))
MergedStates <- inner_join(MainStates, jail_rate_by_state, by = c("Code" = "state"))

g <- ggplot() +
  geom_polygon( data=MergedStates, 
                aes(x=long, y=lat, group=group, fill = total_rate), 
                color="white", size = 0.2)  +
  labs(title="Jail rate by state")
g
```

